---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r preamble, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BNPMIXcluster

<!-- badges: start -->
[![](https://img.shields.io/badge/doi-10.1007/s11634--018--0313--6-yellow.svg)](https://doi.org/10.1007/s11634-018-0313-6)
[![](https://img.shields.io/badge/devel%20version-1.3-blue.svg)](https://github.com/christianu7/BNPMIXcluster)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://cran.r-project.org/web/licenses/MIT)
<!-- badges: end -->

The BNPMIXcluster package provides a method for model-based clustering of multivariate data. It is capable of combining different types of variables (continuous, ordinal and nominal) and accommodates for different sampling probabilities in a complex survey design.

The model is based on a location mixture model with a Poisson-Dirichlet process prior on the location parameters of the associated latent variables.

Details of the underlying model is described in Carmona, C., Nieto-Barajas, L. E., Canale, A. (2016) [arXiv:1612.00083](https://arxiv.org/abs/1612.00083).

## Installation

You can install the released version of BNPMIXcluster from [CRAN](https://CRAN.R-project.org/package=BNPMIXcluster) with:

``` r
install.packages("BNPMIXcluster")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("christianu7/BNPMIXcluster")
```
## Examples

```{r sim_study_5_1}
require(BNPMIXcluster)
```

### Simulation study 1     

In this toy example, we evaluate the clustering allocation of our model using synthetic iid data. Details are discussed in section 5.1 of Carmona et al. (2017).

We sample 3-dimensional latent continuous vectors z=(z1, z2, z3) from a
3-components mixture of Gaussians with equal mixing probabilities and mixture components with means mu_1 = (2, 2, 5), mu_2 = (6, 4, 2) and mu_3 = (1, 6, 2), and variances
Sigma_1 = diag(1, 1, 1), Sigma_2 = diag(0.1, 2, 0.1) and Sigma_3 = diag(2, 0.1, 0.1).

In the next figure we show the "True" (generative) cluster allocation using different colors. Such grouping is unknown and we would like to discover it using our model.

```{r sim_study_5_1_data, echo=FALSE, out.width="70%", fig.align="center"}
require(scatterplot3d)
cluster_color <- c(rgb(1,0,0,alpha = 0.5),
                   rgb(0,0,1,alpha = 0.5),
                   rgb(0,0.5,0,alpha = 0.5))
cluster_color <- cluster_color[Z_latent_ex_5_1$cluster]
cluster_pch <- c(19,15,17)[Z_latent_ex_5_1$cluster]
par(mfrow=c(2,2))
par(mar=c(4,5,2,2))

scatterplot3d::scatterplot3d(x=Z_latent_ex_5_1$Z1,y = Z_latent_ex_5_1$Z2, z=Z_latent_ex_5_1$Z3,
              color=cluster_color,pch=cluster_pch,
              xlab="Z1",ylab="Z2",zlab="Z3",
              main="Simulated data in 3 clusters"
              )
par(mar=c(4,5,2,2))
plot(Z_latent_ex_5_1[,c("Z2","Z3")],col=cluster_color,pch=cluster_pch,xlab="Z2",ylab="Z3")
par(mar=c(4,5,2,2))
plot(Z_latent_ex_5_1[,c("Z1","Z3")],col=cluster_color,pch=cluster_pch,xlab="Z1",ylab="Z3")
par(mar=c(4,5,2,2))
plot(Z_latent_ex_5_1[,c("Z1","Z2")],col=cluster_color,pch=cluster_pch,xlab="Z1",ylab="Z2")
```

We test the method by considering different scenarios for the observed data.It is possible to recover the cluster structure even assuming that we can only observe discretized versions of the latent continuous data.

Here we illustrate the most straightforward scenario (I), where we assume that the three continuous variables are directly observable.

```{r sim_study_5_1_scenario}
# Observable data
# Choose scenario: 1, 2, or 3
ex_i <- 1
head( Y_ex_5_1[[ ex_i ]] )
```

We select hyper-parameters that define a prior specification which promotes a small number of groups.

```{r sim_study_5_1_hyperpars}
# Prior specification
# Choose "a", "b" or "c"
param_j <- "c"
meta_param_ex[ param_j, ]
```

The function `MIXclustering` performs MCMC to allocate clusters to each individual.

```{r sim_study_5_1_mcmc}
set.seed(0) 

# Specify the data type that is being provided to the method
var_type_Y_ex_5_1 <- list( c("c","c","c"),
                           c("o","o"),
                           c("o","o","o","c") )

cluster_ex <- MIXclustering( Y = as.matrix(Y_ex_5_1[[ ex_i ]]),
                             var_type=var_type_Y_ex_5_1[[ ex_i ]],
                             
                             n_iter_out=1500,
                             n_burn=200,
                             n_thin=3,
                             
                             alpha = meta_param_ex[ param_j, "alpha" ],
                             d_0_a = meta_param_ex[ param_j, "d_0_a"],
                             d_1_a = meta_param_ex[ param_j, "d_1_a" ],
                             d_0_b = meta_param_ex[ param_j, "d_0_b" ],
                             d_1_b = meta_param_ex[ param_j, "d_1_b" ],
                             eta = meta_param_ex[ param_j, "eta" ],
                             kappa = meta_param_ex[ param_j, "kappa" ],
                             delta = meta_param_ex[ param_j, "delta" ],
                             
                             d_0_z = meta_param_ex[ param_j, "d_0_z" ],
                             d_1_z = meta_param_ex[ param_j, "d_1_z" ],
                             d_0_mu = meta_param_ex[ param_j, "d_0_mu" ],
                             d_1_mu = meta_param_ex[ param_j, "d_1_mu" ] )

```

### Summary of clustering results

```{r sim_study_5_1_summary}
summary(cluster_ex)
```

### Visualizing clustering results.

Int the first two plots we show the results across all the MCMC samples. First, a heatmap showing the joint allocation of individuals to clusters. Second, the number of clusters obtained along iterations of 
the MCMC. 

```{r sim_study_5_1_plot_1, out.width="50%", fig.align="center"}
# Heatmap of group allocation
plot(cluster_ex,type="heatmap")
```

```{r sim_study_5_1_plot_2, out.width="50%", fig.align="center"}
# Number of clusters obtain across the MCMC chain
plot(cluster_ex,type="chain")
```

### Comparison of resulting cluster configurations.

Once the MCMC has converged, each iteration corresponds to a different clustering configuration for each data point. Here we show some configurations of interest.

The jittered scatter plot below compares:
1. X-axis: Labels from configuration that minimizes the distance with average MCMC iterations
2. Y-axis: Labels from configuration that minimizes an Heterogeneity Measure (HM)

The plot confirms that both configurations agree on the allocation of individuals to their corresponding cluster (Notes: jitter added for clarity, label switching of clusters is natural).

```{r sim_study_5_1_plot_3, out.width="50%", fig.align="center"}
plot( x=jitter(cluster_ex$cluster),y=jitter(cluster_ex$clusterHMmin), col="#FF000080", pch=20,
      main=paste("Comparison of two relevant cluster configurations"),
      xlab="minimizes distance to average MCMC grouping", ylab="minimizes Heterogeneity Measure" )
```

We compare the true (generative) cluster allocation to the configuration suggested by the model:
1. X-axis: Original Labels in the generative model.
2. Y-axis: Labels assigned by the clustering method.

```{r sim_study_5_1_plot_4, out.width="50%", fig.align="center"}
plot(x=jitter(Z_latent_ex_5_1$cluster),
     y=jitter(cluster_ex$cluster),
     main=paste("Comparison real configuration with the model results"),
     xlab="True cluster",
     ylab="Model cluster",
     pch=19, col="#FF000080")
```
